The Weakly Nonlinear Magnetorotational Instability

### 2016-09-02 paper/widegap_mri_paper.tex
boundary layer edits

We find the marginal system as a function of the dimensionless parameters. Marginality for our standard MRI system is a hyperplane in $(\reym, Pm, \beta, \Omega_2/\Omega_1, \mathrm{R}_1/\mathrm{R}_2$), but we hold all of these constant except for $\reym$. To analyze the MRI system at marginality, we fix the parameters listed in Table \ref{table:parameters} and determine the critical $\reym$ and $k_z$ by repeatedly solving the linear MRI system to determine the smallest parameter values for which the fastest growing mode is zero. That is, we solve the linear eigenvalue problem for eigenvalues $\sigma = \gamma + i \omega$. Figure \ref{fig:growth_rates} shows linear MRI growth rates $\gamma$ in the $(\reym, k_z)$ plane. For the fiducial standard MRI parameters in Table \ref{table:parameters} we find critical parameters $\reym_c = 3.30$ and $k_c = 0.901$.

@@

The nonlinear terms, detailed in Appendix \ref{app:nonlinear_terms}, represent an interesting departure from the thin-gap theory. The thin-gap nonlinear terms at both second and third orders are linear combinations of Jacobians. The nonlinear terms in the wide-gap case differ from their thin-gap analogues with in the addition of vertical advective terms. These terms derive from the advective derivatives in the momentum and induction equations, but are filtered out in the thin-gap approximation. These advective terms allow the nonlinear contributions at both second and third order (i.e. N$_2$ and N$_3$) not to individually satisfy the boundary conditions on $\Psi$ and $u$. %(I'm just guessing but this may lead to higher torque on inner cylinder -- should calculate)

We examine the behavior of the wide gap MRI system as a function of $Pm$. Figure \ref{fig:coefficients} shows the critical parameters $k_c$ and $\reym$ as a function of $Pm$, as well as the coefficients $b$, $c$, and $d$. The GLE coefficients are remarkably insensitive to $Pm$. From Equation \ref{eq:gle} it is readily apparent that the asymptotic saturation amplitude is $\alpha_{s} = \pm \sqrt{b/c}$, so we conclude that the saturation amplitude of the MRI is only very weakly dependent on $Pm$. Note that because $\reym$ is essentially constant as a function of $Pm$, the saturation amplitude is equivalently insensitive to $\reye^{-1}$. This is in stark contrast to the narrow gap behavior of the system. For these same boundary conditions, \citeauthor{Umurhan:2007hs} (\citeyear{Umurhan:2007hs}) find that the narrow gap saturation amplitude scales as $Pm^{2/3}$. They find that this amplitude dependence is driven by the $Pm^{1/3}$ dependence of the linear boundary layer. Boundary layer analysis similarly reveals a $\nu^{1/3}$ dependence for the radial extent of the boundary layer in Taylor Couette flow (\citei{Goodman:2002ix}). Why does the $Pm$ scaling of the boundary layer width not translate to a stronger $Pm$ scaling for the saturation amplitude in the wide gap case? The boundary layer in the wide gap case is strongly localized at the walls, i.e. $r_1$ and $r_2$. Figure \ref{fig:wide_narrow_N31B} shows the structure of the third-order nonlinear term $N_{31}^{(B)}$ as a function of $Pm$ for both the narrow and wide gap standard MRI. $N_{31}$ is the vector that determines the GLE coefficient $c$ (see Appendix \ref{app:basic_equations} for the wide gap case, and \citei{Umurhan:2007hs}, \citei{Clark:2016} for the narrow gap equations). Clearly, the boundary layers scale with $Pm$ in both the wide and narrow gap MRI. However, in the narrow gap case this scaling extends prominently into the center of the channel, whereas for the wide gap case the bulk of the flow is relatively unaffected by the boundary layer scaling.

>>>>

We find the marginal system as a function of the dimensionless parameters. The marginal stability curve for our standard MRI system is a hyperplane in $(\reym, Pm, \beta, \Omega_2/\Omega_1, \mathrm{R}_1/\mathrm{R}_2$), but we hold all of these constant except for $\reym$. To analyze the MRI system at marginality, we fix the parameters listed in Table \ref{table:parameters} and determine the critical $\reym$ and $k_z$ by repeatedly solving the linear MRI system to determine the smallest parameter values for which the fastest growing mode is zero. That is, we solve the linear eigenvalue problem for eigenvalues $\sigma = \gamma + i \omega$. Figure \ref{fig:growth_rates} shows linear MRI growth rates $\gamma$ in the $(\reym, k_z)$ plane. For the fiducial standard MRI parameters in Table \ref{table:parameters} we find critical parameters $\reym_c = 3.30$ and $k_c = 0.901$.

@@

The form of the nonlinear terms, detailed in Appendix \ref{app:nonlinear_terms}, represent a departure from the thin-gap theory. The narrow gap nonlinear terms at both second and third orders are linear combinations of Jacobians. The nonlinear terms in the wide-gap case differ from their thin-gap analogues with the addition of vertical advective terms. These terms derive from the advective derivatives in the momentum and induction equations, but are filtered out in the thin-gap approximation. The nonlinear terms ultimately determine the dependence of the saturation amplitude on $Pm$, as described below. %These advective terms allow the nonlinear contributions at both second and third order (i.e. N$_2$ and N$_3$) not to individually satisfy the boundary conditions on $\Psi$ and $u$. %(I'm just guessing but this may lead to higher torque on inner cylinder -- should calculate)

We examine the behavior of the wide gap MRI system as a function of $Pm$. Figure \ref{fig:coefficients} shows the critical parameters $k_c$ and $\reym$ as a function of $Pm$, as well as the GLE linear coefficient $b$ and the diffusion coefficient $d$. The GLE coefficients are remarkably insensitive to $Pm$. From Equation \ref{eq:gle} it is readily apparent that the asymptotic saturation amplitude is $\alpha_{s} = \pm \sqrt{b/c}$, so we conclude that the saturation amplitude of the MRI is only very weakly dependent on $Pm$. Note that because $\reym$ is essentially constant as a function of $Pm$, the saturation amplitude is equivalently insensitive to $\reye^{-1}$. This is in stark contrast to the narrow gap behavior of the system. For these same boundary conditions, \citeauthor{Umurhan:2007hs} (\citeyear{Umurhan:2007hs}) find that the narrow gap saturation amplitude scales as $Pm^{2/3}$. They find that this amplitude dependence is driven by the $Pm^{1/3}$ dependence of the linear boundary layer. Boundary layer analysis similarly reveals a $\nu^{1/3}$ dependence for the radial extent of the boundary layer in Taylor Couette flow (\citei{Goodman:2002ix}). %Why does the $Pm$ scaling of the boundary layer width not translate to a stronger $Pm$ scaling for the saturation amplitude in the wide gap case? The boundary layer in the wide gap case is strongly localized at the walls, i.e. $r_1$ and $r_2$. 
Figure \ref{fig:wide_narrow_N31B} shows the structure of the third-order nonlinear term $N_{31}^{(A)}$ as a function of $Pm$ for both the narrow and wide gap standard MRI. $N_{31}$ is the vector that determines the GLE coefficient $c$, and thus the scaling of the saturation amplitude because of the insensitivity of $b$ to $Pm$ (see Appendix \ref{app:basic_equations} for the wide gap case, and \citei{Umurhan:2007hs}, \citei{Clark:2016} for the narrow gap equations). Clearly, the width of the boundary layers scales with $Pm$ in both the wide and narrow gap MRI. This translates to a steeper saturation amplitude $Pm$ dependence in the wide gap case.%However, in the narrow gap case this scaling extends prominently into the center of the channel, whereas for the wide gap case the bulk of the flow is relatively unaffected by the boundary layer scaling.
